Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach

### 2021-03-18 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Revised the code to aggregate accessibility levels by income quintiles

```{r med-conventional, include=FALSE}
results_5min_conventional <- ttm_walk %>%
  dplyr::filter(hub_type == "Conventional" &
                  hub_status == "Active") %>%
  b2sfca(threshold = 5)
```

```{r med-all, include=FALSE}
results_5min_all <- ttm_walk %>%
  dplyr::filter(hub_status == "Active") %>%
  b2sfca(threshold = 5)
```

```{r, med-sum-conventional, echo=FALSE}
sum(results_5min_conventional$los$los)
sum(results_5min_conventional$accessibility$accessibility)
```

@@

@@

```{r, include=FALSE}
# drop geometry
test <- accessibility %>% st_drop_geometry()

# join accessibility to population cells
test <- test %>% left_join(population_50x50, by = "UID")
```

```{r, include=FALSE}
test0 <- population_50x50 %>%
  st_join(hamilton_da_2016 %>% dplyr::select(DA))
```

```{r, include=FALSE}
test <- test0 %>% left_join(test, by = "UID")
```

```{r, include=FALSE}
# object is a combination of accessibility, population_50x50, with DAs
test <- test %>% drop_na()
```

```{r, include=FALSE}
test <- subset(test, select = -c(1)) %>% dplyr::select(-geometry.y)
```

```{r, include=FALSE}
accessibility <- test
```

```{r, include=FALSE}
min_all <- accessibility %>%
            dplyr::filter(threshold == "3 min") %>% 
            dplyr::filter(system == "All stations")

min_con <- accessibility %>%
            dplyr::filter(threshold == "3 min") %>%
            dplyr::filter(system == "Conventional")

avg_all <- accessibility %>%
            dplyr::filter(threshold == "5 min") %>% 
            dplyr::filter(system == "All stations")

avg_con <- accessibility %>%
            dplyr::filter(threshold == "5 min") %>%
            dplyr::filter(system == "Conventional")

max_all <- accessibility %>%
            dplyr::filter(threshold == "10 min") %>% 
            dplyr::filter(system == "All stations")

max_con <- accessibility %>%
            dplyr::filter(threshold == "10 min") %>%
            dplyr::filter(system == "Conventional")

ext_all <- accessibility %>%
            dplyr::filter(threshold == "15 min") %>% 
            dplyr::filter(system == "All stations")

ext_con <- accessibility %>%
            dplyr::filter(threshold == "15 min") %>%
            dplyr::filter(system == "Conventional")

```

```{r, include=FALSE}
# aggregate population and accessibility by DA
min_all <- min_all %>%
           group_by(DA) %>%
             summarize(population = sum(population.y),
             accessibility = sum(accessibility)) %>%
           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))

min_con <- min_con %>%
           group_by(DA) %>%
             summarize(population = sum(population.y),
             accessibility = sum(accessibility)) %>%
           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))

avg_all <- avg_all %>%
           group_by(DA) %>%
             summarize(population = sum(population.y),
             accessibility = sum(accessibility)) %>%
           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))

avg_con <- avg_con %>%
           group_by(DA) %>%
             summarize(population = sum(population.y),
             accessibility = sum(accessibility)) %>%
           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))

max_all <- max_all %>%
           group_by(DA) %>%
             summarize(population = sum(population.y),
             accessibility = sum(accessibility)) %>%
           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))

max_con <- max_con %>%
           group_by(DA) %>%
             summarize(population = sum(population.y),
             accessibility = sum(accessibility)) %>%
           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))

ext_all <- ext_all %>%
           group_by(DA) %>%
             summarize(population = sum(population.y),
             accessibility = sum(accessibility)) %>%
           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))

ext_con <- ext_con %>%
           group_by(DA) %>%
             summarize(population = sum(population.y),
             accessibility = sum(accessibility)) %>%
           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
```

```{r}
min_all <- min_all %>%
  st_drop_geometry() %>%
  left_join(hamilton_da_2016 %>% 
              dplyr::select(DA, geometry),
            by = "DA") %>%
  st_as_sf()

min_con <- min_con %>%
  st_drop_geometry() %>%
  left_join(hamilton_da_2016 %>% 
              dplyr::select(DA, geometry),
            by = "DA") %>%
  st_as_sf()

avg_all <- avg_all %>%
  st_drop_geometry() %>%
  left_join(hamilton_da_2016 %>% 
              dplyr::select(DA, geometry),
            by = "DA") %>%
  st_as_sf()

avg_con <- avg_con %>%
  st_drop_geometry() %>%
  left_join(hamilton_da_2016 %>% 
              dplyr::select(DA, geometry),
            by = "DA") %>%
  st_as_sf()

max_all <- max_all %>%
  st_drop_geometry() %>%
  left_join(hamilton_da_2016 %>% 
              dplyr::select(DA, geometry),
            by = "DA") %>%
  st_as_sf()

max_con <- max_con %>%
  st_drop_geometry() %>%
  left_join(hamilton_da_2016 %>% 
              dplyr::select(DA, geometry),
            by = "DA") %>%
  st_as_sf()

ext_all <- ext_all %>%
  st_drop_geometry() %>%
  left_join(hamilton_da_2016 %>% 
              dplyr::select(DA, geometry),
            by = "DA") %>%
  st_as_sf()

ext_con <- ext_con %>%
  st_drop_geometry() %>%
  left_join(hamilton_da_2016 %>% 
              dplyr::select(DA, geometry),
            by = "DA") %>%
  st_as_sf()
```

```{r}
data_min_all <- bi_class(min_all,
                 x = accessibility, 
                 y = median_total_income, 
                 style = "quantile", 
                 dim = 3) %>%
  st_as_sf()

data_min_con <- bi_class(min_con,
                 x = accessibility, 
                 y = median_total_income, 
                 style = "quantile", 
                 dim = 3) %>%
  st_as_sf()

data_avg_all <- bi_class(avg_all,
                 x = accessibility, 
                 y = median_total_income, 
                 style = "quantile", 
                 dim = 3) %>%
  st_as_sf()

data_avg_con <- bi_class(avg_con,
                 x = accessibility, 
                 y = median_total_income, 
                 style = "quantile", 
                 dim = 3) %>%
  st_as_sf()

data_max_all <- bi_class(max_all,
                 x = accessibility, 
                 y = median_total_income, 
                 style = "quantile", 
                 dim = 3) %>%
  st_as_sf()

data_max_con <- bi_class(max_con,
                 x = accessibility, 
                 y = median_total_income, 
                 style = "quantile", 
                 dim = 3) %>%
  st_as_sf()

data_ext_all <- bi_class(ext_all,
                 x = accessibility, 
                 y = median_total_income, 
                 style = "quantile", 
                 dim = 3) %>%
  st_as_sf()

data_ext_con <- bi_class(ext_con,
                 x = accessibility, 
                 y = median_total_income, 
                 style = "quantile", 
                 dim = 3) %>%
  st_as_sf()
```
```{r figure-10, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) varying by median total income in the current system with equity stations."}
map_min_all <- ggplot() +
  geom_sf(data = data_min_all, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "DkBlue", dim = 3) +
  bi_theme()

>>>>

```{r med-conventional, include=FALSE}
results_6min_conventional <- ttm_walk %>%
  dplyr::filter(hub_type == "Conventional" &
                  hub_status == "Active") %>%
  b2sfca(threshold = 6)
```

```{r med-all, include=FALSE}
results_6min_all <- ttm_walk %>%
  dplyr::filter(hub_status == "Active") %>%
  b2sfca(threshold = 6)
```

```{r, med-sum-conventional, echo=FALSE}
sum(results_6min_conventional$los$los)
sum(results_6min_conventional$accessibility$accessibility)
```

@@

```{r include=FALSE}
# Testing different thresholds to check how accessibility changes
accessibility_test <- data.frame(threshold = c(3:15), 
                                 accessibility = numeric(13))
for(i in 1:13){
  results_test <- ttm_walk %>%
  dplyr::filter(hub_type == "Conventional" &
                  hub_status == "Active") %>%
  b2sfca(threshold = i +2)
  accessibility_test$accessibility[i] <- sum(results_test$accessibility$accessibility)
}

ggplot(accessibility_test, 
       aes(x = threshold, y = accessibility)) +
  geom_line()
```

@@

```{r, include=FALSE}
# drop geometry
test <- accessibility

# join accessibility to population cells
test <- test %>% st_join(population_50x50 %>% dplyr::select(-UID)) %>%
  rename(population50x50 = population)
```

```{r}
test <- test %>%
  st_join(hamilton_da_2016)
```

```{r}
agg_accessibility_da <- test %>%
  group_by(DA, 
           system,
           threshold) %>%
  summarize(accessibility = sum(accessibility),
            median_total_income = first(median_total_income), 
            .groups = "drop") %>%
  st_drop_geometry() %>%
  left_join(hamilton_da_2016 %>% dplyr::select(DA, geometry),
            by = c("DA" = "DA")) %>%
  st_as_sf()
```

```{r}
data_map <- agg_accessibility_da %>%
            dplyr::filter(threshold == "3 min", 
                          system == "All stations") %>%
  bi_class(x = accessibility, 
           y = median_total_income, 
           style = "quantile", 
           dim = 3) %>%
  st_as_sf()

map_min_all <- ggplot() +
  geom_sf(data = data_map,
          mapping = aes(fill = bi_class), 
          color = "white", 
          size = 0.1, 
          show.legend = FALSE) +
  bi_scale_fill(pal = "DkBlue", 
                dim = 3) +
  bi_theme()

Coming back to the question of accessibility by income:
```{r}
# Find the income quintiles of the population served by SoBi
iq <- hamilton_da_2016 %>% 
  dplyr::filter(sobi_service == "Yes") %>%
  pull(median_total_income) %>%
  quantile(c(0, 0.2, 0.4, 0.6, 0.8, 1), na.rm = TRUE)
```

```{r}
# Add a variable to identify the income 
agg_accessibility_income <- test %>% 
  dplyr::filter(system == "All stations", threshold == "10 min") %>%
  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
                                     median_total_income > iq[5] ~ "Top 20%"))

inc_stats_All_10 <- agg_accessibility_income %>%
  st_drop_geometry() %>%
  dplyr::filter(system == "All stations", threshold == "10 min") %>%
  group_by(income_quintile) %>%
  summarize(accessibility = sum(accessibility),
            population = sum(population50x50))
```

```{r}
# Add a variable to identify the income 
agg_accessibility_income <- test %>% 
  dplyr::filter(system == "Conventional", threshold == "10 min") %>%
  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
                                     median_total_income > iq[5] ~ "Top 20%"))

inc_stats_Conv_10 <- agg_accessibility_income %>%
  st_drop_geometry() %>%
  dplyr::filter(system == "Conventional", threshold == "10 min") %>%
  group_by(income_quintile) %>%
  summarize(accessibility = sum(accessibility),
            population = sum(population50x50))
```

```{r}
rbind(inc_stats_All_10,  
      inc_stats_Conv_10)
```

```{r figure-10, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) varying by median total income in the current system with equity stations."}
map_min_all <- ggplot() +
  geom_sf(data = data_min_all, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "DkBlue", dim = 3) +
  bi_theme()
